Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-02-11 WholePlotTrajectories.Rmd
intro, dicussion

Examine post-disturbance trajectories requires a set of diversity metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
Specifically, diversity metrics should assess taxonomic richness, evenness and composition, in order to highlight the changes the community abundance distribution and in the species fostered by the disturbance. 
Analysis would as well include a functional approach accounting for species biological attributes and ecological strategy.
The trait-based approach directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we (i) elucidated community taxonomic and functional resilience, (ii) discussed the scope of the IDH and the underlying ecological processes, and (iii) examined community recovery time.

@@

## A cyclic recovery of community composition

The hump-shaped trajectories of community composition returned towards pre-disturbance state after 30 years, highlighting community resilience in terms of taxonomic and functional composition.

Before disturbance, there were already taxonomic differences among local communities, as revealed by the distinct starting points on the NMDS axis 2.
These initial differences were maintained throughout the trajectories following disturbance, so community initial state influenced the trajectories following disturbance more than commonly thought.
The initial composition indeed partly determined the pool of species recruited, and hence shaped the composition trajectories and drove them towards the recovery of the initial composition.
This taxonomic recovery, although it was not fully achieved after 30 years, suggested the resilience of community taxonomic composition and the maintenance of community initial differences [@Folke2006].
Species not belonging to the pre-disturbance community were then rarely recruited, probably because of the common dispersal limitations among tropical tree species [@Svenning2005].

Functional composition trajectories were, conversely, highly similar among disturbed communities.
As the composition of pre-disturbance surviving trees is representative of the initial community [@Herault2018], changes in functional composition relied upon the recruitment of species or functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches vacated by the disturbance, with a high availability of light, space and nutrients.
This recruitment of pioneers similarly changed the functional composition of all communities, whenever the disturbance intensity.
The functional composition went towards more resource-acquisitive strategies, translated in the functional plan (Fig. \@ref(fig:NMDSplans)) by a displacement on the right along the first axis [@Westoby1998; @Wright2004; @Reich2014].
Thereafter, the pioneers first recruited were progressively excluded by long-lived, more resistant and shade-tolerant species.
The commpunity functional composition then returned towards more resource-conservative strategies, suggesting the recovery of the initial community composition and translated in the functional plan by a displacement left along the first axis and upward along the second axis (Fig. \@ref(fig:NMDSplans)).

In the light of these taxonomic and functional composition trajectories, community response to disturbance proved simultaneously deterministic and stochastic at different levels of community organization.
Taxonomic trajectories were shaped by dispersal limitations maintaining the stochastic differences among communities due to historical contingency. 
Functional trajectories were shaped by trait-based processes resulting the the convergence of community functional composition [@Fukami2005].

@@

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold for which community response corresponded to a maximized richness and a resilience of the evenness. 
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pionners, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they besides temporarily balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The richness trajectories following intense disturbance then decreased, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

Observed trajectories supported the IDH in highlighting a threshold disturbance intensity, expressed as percentage of AGB removed, marking a change in post-disturbance trajectories.
Taxonomic trajectories besides highlighted, below the intensity threshold, an intermediate time after when the taxonomic evenness was maximized.
This time corresponded to the intermediate between the generation time of pionneers, first to dominate the post-disturbance recruitment, and the time for competitive exclusion, leading to the dominance of late-successional species.
As already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998],
taxonomic evenness followed humped-shaped trajectories with a peak when the recruitment was balanced between pioneers and late-successional species.

Regarding community functional trajectories in contrast, there was neither intermediate disturbance intensity nor time to disturbance, which dismissed the IDH.
The functional response according to disturbance intensity did not display the characteristic hump-shaped trajectories.
Whenever the disturbance intensity, both community functional richness and evenness increased with the recruitment of pioneers that were functionally highly different from the species of the pre-disturbance community [@Denslow1980;@Molino2001].
Above the intensity threshold however, for the most intense disturbance, community functional richness and evenness started to decrease after 15 to 20 years. 
The vacated environmental space was then occupied by short-lived species which prevented the establishment of other species. 
The decline of these short-lived pioneers decreased the functional richness and evenness, but we suggest that the establishment of long-lasting pioneers will follow and the taxonomic and functional trajectories will catch up with those observed for intermediate disturbance [@Walker2009].

The IDH then translated into community taxonomic response, with markedly different trajectories below and above an intensity threshold, and with an intermediate time to disturbance when the intensity remained below the threshold. 
It was however not tangible regarding the functional response, with functional trajectories decoupled from taxonmic ones.

@@

The decoupling between taxonomic and functional trajectories was explained by a decrease in the functional redundancy within the pre-disturbance functional space, due to the loss of species following disturbance.
The recruitment of pioneers, functionally different from the pre-disturbance functional composition, did not compensate the decrease of functional redundancy in the first place.
Progressively though, the functional redundancy was restored through the replacement of the first established species by more competitive, long-lived pioneers or late-successional species that were functionally closer to the pre-disturbance community.
This replacement followed the lottery recruitment rules, implying a recruitment easy for the first recruited species but increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
The functional redundancy then relied upon stochastic processes and its recovery, depending on the recruitment of infrequent species constrained by the increasing competition, was increasingly slow and difficult to anticipate [@Elmqvist2003;@Diaz2005].

The impact of disturbance on community functional redundancy meant a lower resilience of the pre-disturbance communities, with higher chances to see the persistence of disturbance-specific species at the expense of late-successional ones [@Haddad2008].
Besides, the stochastic recovery of infrequent species increases the risks to lose keystone species, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005].
Apart from the functional characteristics considered here, infrequent species might indeed have unique functions in the ecosystem or be a key for some fauna [@Schleuning2016].

>>>>

Fully analyse post-disturbance trajectories requires a set of metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
The analysis should first consider community composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006] and assesses the different species fostered by the disturbance.
Diversity metrics encompassing taxonomic richness and evenness would assess the changes community abundance distribution. 
A functional approach accounting for species biological attributes and ecological strategy, would directly link community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, (i) from community trajectories we examined their taxonomic and functional resilience, (ii) discussed the scope of the IDH and the underlying ecological processes, and (iii) analyzed community time to recovery.

@@

## Decoupled taxonomic and functional trajectories

Before disturbance, the different communities had different taxonomic compositions, visualized by their distinct starting points on the NMDS axis 2.
These initial differences were maintained all along the 30 years following disturbance, with post-disturbance trajectories corresponding to a displacement on the NMDS axis 1.
The taxonomic compositional changes following disturbance were similar among communities, which may correspond to the recruitment of a group of pioneers common to all communities whenever their initial differences and the disturbance intensity [@Denslow2000;@Bongers2009].
Taxonomic trajectories continued with a recovery of the initial composition which, althought not fully achieved after 30 years, suggested the community taxonomic resilience and the maintenance of community initial differences [@Folke2006].
The  return towards initial taxonomic composition showed that species not belonging to the pre-disturbance community were rarely recruited, probably because of the common dispersal limitations among tropical tree species [@Svenning2005].

Regarding functional composition, initial communities were similar and trajectories were confounded in the functional space.
Following disturbance, the functional composition of pre-disturbance surviving trees is belived to be the same as this of the initial community [@Herault2018].
Post-disturbance changes in functional composition then relied upon the recruitment of species or functional types previously infrequent or absent.
These species were probably competitive pioneers becoming dominant as the most efficient to benefit from the light, space and nutrients made available by disturbance.
This recruitment of pioneers drove functional composition trajectories similar for all plots and disturbance intensity.
The composition first went towards more resource-acquisitive strategies, which was translated by a displacement on the right along the first axis in the functional plan [@Westoby1998; @Wright2004; @Reich2014].
Thereafter, the pioneers recruited primarily were progressively excluded by long-lived, more resistant and shade-tolerant species.
The community functional composition then returned towards more resource-conservative strategies, suggesting the recovery of the initial community composition and translated in the functional plan by a displacement left along the first axis and upward along the second axis (Fig. \@ref(fig:NMDSplans)).

Both taxonomic and functional composition trajectories initiated a return towards pre-disturbance state after 30 years, which highlighted community taxonomic and functional resilience. 
There was however a decoupling between taxonomic and functional composition trajectories as taxonomic trajectories maintained the initial differences among communities, while functional trajectories where similar and convergent in the functional space [@Fukami2005].

@@

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold for which the taxonomic richness was maximized and the taxonomic evenness was resilient. 
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pionners, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness then decreased following intense disturbance, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity besides highlighted an intermediate time after disturbance for which the taxonomic evenness was maximized.
As already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998],
taxonomic evenness followed hump-shaped trajectories with a peak when the recruitment was balanced between pioneers and late-successional species. 
This time then corresponded to the match between the generation time of pionneers, first to dominate the post-disturbance recruitment, and the time for competitive exclusion to emerge, which leads to the dominance of late-successional species.

Regarding community functional trajectories in contrast, there was neither intermediate disturbance intensity nor intermediate time after disturbance, which dismissed the IDH.
Neither functional richness nor functional evenness displayed a humped-shaped pattern, but both rather increased with the recruitment of pioneers that were functionally highly different from the pre-disturbance community [@Denslow1980;@Molino2001].
Above the intensity threshold however, for the most intense disturbance, community functional richness and evenness started to decrease after 15 to 20 years. 
The right after disturbance, short-lived species benefited from the resources made available and prevented the establishment of other species. 
The decline of these short-lived pioneers decreased the functional richness and evenness, but we suggest that the establishment of long-lasting pioneers will follow and that the trajectories will catch up with those observed for intermediate disturbance [@Walker2009].

The IDH then translated into community taxonomic response, with post-disturbance trajectories markedly different below and above an intensity threshold.
The IDH was also valid at an intermediate time after disturbance, when the recruitment of early- and late-successional species were balanced. 
Functional trajectories, however, remained decoupled from taxonomic ones and dismissed the IDH.

@@

The decoupling between taxonomic and functional trajectories was explained by a decrease in the functional redundancy within the pre-disturbance functional space, due to the loss of species following disturbance.
The recruitment of pioneers, functionally different from the pre-disturbance functional composition, did not compensate the decrease of functional redundancy in the first place.
Progressively though, the functional redundancy was restored through the replacement of the first established species by more competitive, long-lived pioneers or late-successional species that were functionally closer to the pre-disturbance community.
This replacement followed the lottery recruitment rules, implying an recruitment easy for the first species but becoming increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
The recovery of the functional redundancy then relied upon the random processe of species recruitment and was increasingly slow and difficult to anticipate [@Elmqvist2003;@Diaz2005].

The impact of disturbance on community functional redundancy meant a lower resilience of the disturbed communities, with higher chances to see the settltement of pioneers and early-successional species at the expense of late-successional ones [@Haddad2008].
Besides, the random recovery of infrequent species increases the risks to lose keystone species, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005].
Infrequent species might indeed have unique functional characteristics, apart from those considered here, in the ecosystem or be a key for some fauna [@Schleuning2016].

### 2019-02-11 WholePlotTrajectories.tex
intro, dicussion

Examine post-disturbance trajectories requires a set of diversity
metrics, to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. Specifically, diversity
metrics should assess taxonomic richness, evenness and composition, in
order to highlight the changes the community abundance distribution and
in the species fostered by the disturbance. Analysis would as well
include a functional approach accounting for species biological
attributes and ecological strategy. The trait-based approach directly
links community diversity, composition and redundancy to ecosystem
functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits that represent species ecological
strategy and determine how they respond to changing conditions
\citep{Diaz2005}. Specifically in tropical forests, the functional
approach revealed the post-disturbance deterministic processes entailing
a shift from a dominance of ``conservative'' slow-growing species
dealing with scarce resources, to a dominance of ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{Rees2001, Reich2014, Herault2011}. This shift is translated into
the trajectories of average community value of key functional traits
related to resource acquisition (leaf and stem traits) and life-history
strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
resilience, (ii) discussed the scope of the IDH and the underlying
ecological processes, and (iii) examined community recovery time.

@@

\subsection{A cyclic recovery of community
composition}\label{a-cyclic-recovery-of-community-composition}

The hump-shaped trajectories of community composition returned towards
pre-disturbance state after 30 years, highlighting community resilience
in terms of taxonomic and functional composition.

Before disturbance, there were already taxonomic differences among local
communities, as revealed by the distinct starting points on the NMDS
axis 2. These initial differences were maintained throughout the
trajectories following disturbance, so community initial state
influenced the trajectories following disturbance more than commonly
thought. The initial composition indeed partly determined the pool of
species recruited, and hence shaped the composition trajectories and
drove them towards the recovery of the initial composition. This
taxonomic recovery, although it was not fully achieved after 30 years,
suggested the resilience of community taxonomic composition and the
maintenance of community initial differences \citep{Folke2006}. Species
not belonging to the pre-disturbance community were then rarely
recruited, probably because of the common dispersal limitations among
tropical tree species \citep{Svenning2005}.

Functional composition trajectories were, conversely, highly similar
among disturbed communities. As the composition of pre-disturbance
surviving trees is representative of the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types that were infrequent or
absent before disturbance. Competitive pioneers became dominant in
filling the environmental niches vacated by the disturbance, with a high
availability of light, space and nutrients. This recruitment of pioneers
similarly changed the functional composition of all communities,
whenever the disturbance intensity. The functional composition went
towards more resource-acquisitive strategies, translated in the
functional plan (Fig. \ref{fig:NMDSplans}) by a displacement on the
right along the first axis \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter, the pioneers first recruited were progressively excluded by
long-lived, more resistant and shade-tolerant species. The commpunity
functional composition then returned towards more resource-conservative
strategies, suggesting the recovery of the initial community composition
and translated in the functional plan by a displacement left along the
first axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

In the light of these taxonomic and functional composition trajectories,
community response to disturbance proved simultaneously deterministic
and stochastic at different levels of community organization. Taxonomic
trajectories were shaped by dispersal limitations maintaining the
stochastic differences among communities due to historical contingency.
Functional trajectories were shaped by trait-based processes resulting
the the convergence of community functional composition
\citep{Fukami2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which community response corresponded to a maximized richness and a
resilience of the evenness. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pionners, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they besides temporarily balanced
the usual hyper-dominance of tropical forests, hence temporarily
increasing the taxonomic evenness \citep{Baraloto2012a}. Above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The richness trajectories
following intense disturbance then decreased, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

Observed trajectories supported the IDH in highlighting a threshold
disturbance intensity, expressed as percentage of AGB removed, marking a
change in post-disturbance trajectories. Taxonomic trajectories besides
highlighted, below the intensity threshold, an intermediate time after
when the taxonomic evenness was maximized. This time corresponded to the
intermediate between the generation time of pionneers, first to dominate
the post-disturbance recruitment, and the time for competitive
exclusion, leading to the dominance of late-successional species. As
already observed in the Guiana shield \citep{Baraloto2012a} and in
Borenan tropical forests \citep{Cannon1998}, taxonomic evenness followed
humped-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species.

Regarding community functional trajectories in contrast, there was
neither intermediate disturbance intensity nor time to disturbance,
which dismissed the IDH. The functional response according to
disturbance intensity did not display the characteristic hump-shaped
trajectories. Whenever the disturbance intensity, both community
functional richness and evenness increased with the recruitment of
pioneers that were functionally highly different from the species of the
pre-disturbance community \citep{Denslow1980, Molino2001}. Above the
intensity threshold however, for the most intense disturbance, community
functional richness and evenness started to decrease after 15 to 20
years. The vacated environmental space was then occupied by short-lived
species which prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and the taxonomic and functional trajectories will catch up
with those observed for intermediate disturbance \citep{Walker2009}.

The IDH then translated into community taxonomic response, with markedly
different trajectories below and above an intensity threshold, and with
an intermediate time to disturbance when the intensity remained below
the threshold. It was however not tangible regarding the functional
response, with functional trajectories decoupled from taxonmic ones.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying a recruitment easy for the first recruited
species but increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The functional redundancy then relied
upon stochastic processes and its recovery, depending on the recruitment
of infrequent species constrained by the increasing competition, was
increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

The impact of disturbance on community functional redundancy meant a
lower resilience of the pre-disturbance communities, with higher chances
to see the persistence of disturbance-specific species at the expense of
late-successional ones \citep{Haddad2008}. Besides, the stochastic
recovery of infrequent species increases the risks to lose keystone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Apart from the functional
characteristics considered here, infrequent species might indeed have
unique functions in the ecosystem or be a key for some fauna
\citep{Schleuning2016}.

>>>>

Fully analyse post-disturbance trajectories requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006} and assesses the different species
fostered by the disturbance. Diversity metrics encompassing taxonomic
richness and evenness would assess the changes community abundance
distribution. A functional approach accounting for species biological
attributes and ecological strategy, would directly link community
diversity, composition and redundancy to ecosystem functioning and
environmental constraints \citep{Violle2007b, Baraloto2012a}. In that
respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) from community trajectories we examined their
taxonomic and functional resilience, (ii) discussed the scope of the IDH
and the underlying ecological processes, and (iii) analyzed community
time to recovery.

@@

\subsection{Decoupled taxonomic and functional
trajectories}\label{decoupled-taxonomic-and-functional-trajectories}

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with post-disturbance trajectories corresponding
to a displacement on the NMDS axis 1. The taxonomic compositional
changes following disturbance were similar among communities, which may
correspond to the recruitment of a group of pioneers common to all
communities whenever their initial differences and the disturbance
intensity \citep{Denslow2000, Bongers2009}. Taxonomic trajectories
continued with a recovery of the initial composition which, althought
not fully achieved after 30 years, suggested the community taxonomic
resilience and the maintenance of community initial differences
\citep{Folke2006}. The return towards initial taxonomic composition
showed that species not belonging to the pre-disturbance community were
rarely recruited, probably because of the common dispersal limitations
among tropical tree species \citep{Svenning2005}.

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is belived to be the same as this of the initial community
\citep{Herault2018}. Post-disturbance changes in functional composition
then relied upon the recruitment of species or functional types
previously infrequent or absent. These species were probably competitive
pioneers becoming dominant as the most efficient to benefit from the
light, space and nutrients made available by disturbance. This
recruitment of pioneers drove functional composition trajectories
similar for all plots and disturbance intensity. The composition first
went towards more resource-acquisitive strategies, which was translated
by a displacement on the right along the first axis in the functional
plan \citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the
pioneers recruited primarily were progressively excluded by long-lived,
more resistant and shade-tolerant species. The community functional
composition then returned towards more resource-conservative strategies,
suggesting the recovery of the initial community composition and
translated in the functional plan by a displacement left along the first
axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
community taxonomic and functional resilience. There was however a
decoupling between taxonomic and functional composition trajectories as
taxonomic trajectories maintained the initial differences among
communities, while functional trajectories where similar and convergent
in the functional space \citep{Fukami2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness was resilient. The disturbance intensity determined the balance
in the community between pre-disturbance surviving trees, and trees
recruited afterward. Below a 20\%-25\% AGB loss, the trees surviving
after disturbance remained numerous enough to maintain the high
taxonomic richness of the pre-disturbance community \citep{Bongers2009}.
The recruitment of pionners, infrequent or absent before disturbance,
then increased the taxonomic richness all the more so that the
disturbance was intense \citep{Martin2015, Chaudhary2016}. As these
pioneers became more dominant they balanced the usual hyper-dominance of
tropical forests, hence temporarily increasing the taxonomic evenness
\citep{Baraloto2012a}. Above the intensity threshold, the taxonomic
richness of surviving trees was too low to be offset by the recruitment
of pioneers. In the Guiana shield indeed, the pool of true pioneers
specifically recruited after disturbance is restricted to a few common
genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance for which the
taxonomic evenness was maximized. As already observed in the Guiana
shield \citep{Baraloto2012a} and in Borenan tropical forests
\citep{Cannon1998}, taxonomic evenness followed hump-shaped trajectories
with a peak when the recruitment was balanced between pioneers and
late-successional species. This time then corresponded to the match
between the generation time of pionneers, first to dominate the
post-disturbance recruitment, and the time for competitive exclusion to
emerge, which leads to the dominance of late-successional species.

Regarding community functional trajectories in contrast, there was
neither intermediate disturbance intensity nor intermediate time after
disturbance, which dismissed the IDH. Neither functional richness nor
functional evenness displayed a humped-shaped pattern, but both rather
increased with the recruitment of pioneers that were functionally highly
different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Above the intensity threshold however,
for the most intense disturbance, community functional richness and
evenness started to decrease after 15 to 20 years. The right after
disturbance, short-lived species benefited from the resources made
available and prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and that the trajectories will catch up with those observed
for intermediate disturbance \citep{Walker2009}.

The IDH then translated into community taxonomic response, with
post-disturbance trajectories markedly different below and above an
intensity threshold. The IDH was also valid at an intermediate time
after disturbance, when the recruitment of early- and late-successional
species were balanced. Functional trajectories, however, remained
decoupled from taxonomic ones and dismissed the IDH.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an recruitment easy for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random processe of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

The impact of disturbance on community functional redundancy meant a
lower resilience of the disturbed communities, with higher chances to
see the settltement of pioneers and early-successional species at the
expense of late-successional ones \citep{Haddad2008}. Besides, the
random recovery of infrequent species increases the risks to lose
keystone species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Infrequent species might
indeed have unique functional characteristics, apart from those
considered here, in the ecosystem or be a key for some fauna
\citep{Schleuning2016}.
